Software architectures to integrate workflow engines in science gateways

### 2016-10-17 answers.tex
Updates following Marc's comments. Submitted revision

  The results and discussion reported in our paper are indeed aimed at
  an expert audience of science gateway and workflow engine
  developers. The evaluation metrics have been designed based on our
  long experience with both science gateways and workflow engines, and
  we completely acknowledge that non experts may misinterpret the
  results.

@@

\revised{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

@@

\begin{answer}
  We hope that the explanations provided in this letter and the
  updated text in the paper demonstrate that the metrics and
  definitions are rigorous.
\end{answer}

>>>>

  The results and discussion reported in our paper are indeed aimed at
  an expert audience of science gateway and workflow engine
  developers. The evaluation metrics have been designed based on our
  long experience with both science gateways and workflow engines, and
  we completely acknowledge it may be of limited interest for non
  expert users.

@@

\revised{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science gateways.
  In real systems, such patterns sometimes coexist due to the
  historical and technical context of software projects.}

@@

\begin{answer}
  We hope that the explanations provided in this letter and the
  updated text in the paper demonstrate that the metrics and
  definitions are rigorous. We also hope that the flexibility provided
  by the shared metrics sheet will bring the readers to consider this
  paper to be a valuable tool to stimulate in depth discussions based
  on an initial comparison between abstract architectural designs.
\end{answer}

### 2016-10-17 paper.tex
Updates following Marc's comments. Submitted revision

\correction{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

@@

The architecture of the Catania Science Gateway Framework (CSGF) is
shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
Liferay portlets, i.e., as part of the science gateway when no
specific protocol is used for remote invocation. A specific portlet
called \texttt{mi-parallel-portlet} allows to create job collections,
parametric jobs and split-and-merge types of workflows. The science
gateway also includes a set of services to manage data and tasks
(jobs) on a wide array of grid middleware systems through the SAGA
API.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/CSGF}
\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
  integration). Figure adapted from~\cite{ardizzone2012decide} with
  permission of the first author} \todo{approval still pending}.}
\label{fig:csgf}
\end{figure}

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform (VIP), where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service. The interactions
in the VIP architecture map to the ones defined by our framework in a
straightforward manner, except for interaction \texttt{c$_2$}
(management of workflow data) which, in VIP, is triggered by the
workflow tasks instead of the workflow engine. This does not
contradict the service invocation architecture since the tasks are
generated by the workflow engine.
\begin{figure*}
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
  workflow data transfers are performed by the tasks and embedded in
  their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
adapted from~\cite{GLAT-16}.]{
  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3}), and workflow
  retrieval (\texttt{13} \& \texttt{14}). Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure adapted
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

Figure~\ref{fig:shiwa-architecture} shows the architecture used in the
SHIWA Simulation Platform for nested workflow execution with service
invocation. The parent workflow engine is \texttt{WS-PGRADE/gUSE},
invoked as a service in the Science Gateway (interaction
\texttt{d}). Ten different child engines can be used by nested
workflows, invoked through the \texttt{Submission service}
(interaction \texttt{d}). Each of these engines can submit tasks and
transfer data to a distributed computing infrastructure (\texttt{DCI}
interactions \texttt{b} and \texttt{c$_2$}). User data interactions
(\texttt{c$_1$}) and interactions between the parent engine and the
infrastructure (\texttt{b} and \texttt{c$_2$}) are not represented.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.png}
\caption{Nested workflow execution through the SHIWA Science Gateway. Figure
  adapted from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

\section{\correction{Architecture evaluations}}

@@

\correction{To illustrate how our evaluation framework can be applied
  to help design new systems, let's consider a hypothetical system
  where the workflow engine is split in two parts: (1) a high-level
  part integrated in the science gateway, which submits sub-workflows
  to the infrastructure; (2) a lower-level part, executed on the
  infrastructure, which executes the sub-workflows. This scenario came
  up in the discussion with Reviewer \#2 of this paper, after the
  evaluation framework had already been designed.

>>>>

\correction{This analysis is intended for experts of science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science gateways.
  In real systems, such patterns sometimes coexist due to the
  historical and technical context of software projects.}

@@

The architecture of the Catania Science Gateway Framework (CSGF) is
shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
Liferay portlets, i.e., as part of the science gateway when no
specific protocol is used for remote invocation. A specific portlet
called \texttt{mi-parallel-portlet} allows to create job collections,
parametric jobs and split-and-merge types of workflows. The science
gateway also includes a set of services to manage data and tasks
(jobs) on a wide array of grid middleware systems through the SAGA
API.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/CSGF}
\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
  integration). Figure adapted from~\cite{ardizzone2012decide} with
  permission of the first author} \color{red}({approval still pending})\color{black}.}
\label{fig:csgf}
\end{figure}

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform (VIP), where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service. The interactions
in the VIP architecture map to the ones defined by our framework in a
straightforward manner, except for interaction \texttt{c$_2$}
(management of workflow data) which, in VIP, is triggered by the
workflow tasks instead of the workflow engine. This does not
contradict the service invocation architecture since the tasks are
generated by the workflow engine.
\begin{figure*}
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
\caption{\correction{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
  workflow data transfers are performed by the tasks and embedded in
  their descriptions).}}
\label{fig:vip-architecture}
\end{figure*}

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
adapted from~\cite{GLAT-16}.]{
  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
\label{fig:cbrain-psom-architecture}
}
\caption{\correction{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{$c_1$} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{\correction{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3}), and workflow
  retrieval (\texttt{13} \& \texttt{14}). Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure adapted
  from~\cite{ROGE-13}.}}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

Figure~\ref{fig:shiwa-architecture} shows the architecture used in the
SHIWA Simulation Platform for nested workflow execution with service
invocation. The parent workflow engine is \texttt{WS-PGRADE/gUSE},
invoked as a service in the Science Gateway (interaction
\texttt{d}). Ten different child engines can be used by nested
workflows, invoked through the \texttt{Submission service}
(interaction \texttt{d}). Each of these engines can submit tasks and
transfer data to a distributed computing infrastructure (\texttt{DCI}
interactions \texttt{b} and \texttt{c$_2$}). User data interactions
(\texttt{c$_1$}) and interactions between the parent engine and the
infrastructure (\texttt{b} and \texttt{c$_2$}) are not represented.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.png}
\caption{\correction{Nested workflow execution through the SHIWA Science Gateway. Figure
  adapted from~\cite{terstyanszky2014enabling} with permission of
  the first author.}}
\label{fig:shiwa-architecture}
\end{figure*}

@@

\section{\correction{Architecture evaluation}}

@@

\correction{To illustrate how our evaluation framework can be applied
  to help design new systems, let's consider a hypothetical system
  where the workflow engine is split in two parts: (1) a high-level
  part integrated in the science gateway, which submits sub-workflows
  to the infrastructure; (2) a lower-level part, executed on the
  infrastructure, which executes the sub-workflows. This scenario came
  up after the evaluation framework had already been designed.
